Intersection Types, Sub-formula Property, and the Functional Character of the Lambda Calculus

### 2018-08-09 DenotSem/determ.tex
check

Many of the challenges revolved around the definitions of
$\sqsubseteq$, consistency, and $\sqcup$. Given that I already had
definitions for $\sqsubseteq$ and consistency, the obvious thing to
try was to define $\sqcup$ such that it would be the least upper bound
of $\sqsubseteq$. So I arrived at this partial function:
\begin{align*}
  n \sqcup n &= n \\
  f_1 \sqcup f_2 &= f_1 \cup f_2
\end{align*}
Now suppose we prove the Join Theorem by induction on $e$ and consider
the case for application: $e = (e_1 \app e_2)$. From $v \in \ESEM{e_1
  \app e_2}$ and $v' \in \ESEM{e_1 \app e_2}$ we have
\begin{itemize}
\item $f \in \ESEM{e_1}\rho$, $v_2 \in \ESEM{e_2}\rho$,  
  $v_3 \mapsto v_4 \in f$, $v_3 \sqsubseteq v_2$, and $v \sqsubseteq v_4$
  for some $f, v_2, v_3, v_4$.
\item $f' \in \ESEM{e_2}\rho'$, $v'_2 \in \ESEM{e_2}\rho'$,   
  $v'_3 \mapsto v'_4 \in f$, $v'_3 \sqsubseteq v'_2$, and $v' \sqsubseteq v'_4$
  for some $f', v'_2, v'_3, v'_4$.
\end{itemize}
By the induction hypothesis we have $f \sqcup f' \in \ESEM{e_1}$
and $v_2 \sqcup v'_2 \in \ESEM{e_2}$.
We need to show that 
\[
   v''_3 \mapsto v''_4 \in f \sqcup f' 
   \qquad
   v''_3 \sqsubseteq v_2 \sqcup v'_2
   \qquad
   v \sqcup v' \sqsubseteq v''_4
\]
But here we have a problem. Given our definition of $\sqcup$ in terms
of set union, there won't necissarily be a single entry in $f \sqcup
f'$ that combines the information from both $v_3 \mapsto v_4$ and
$v'_3 \mapsto v'_4$. After all, $f \sqcup f'$ contains all the entries
of $f$ and all the entries of $f'$, but the set union operation does
not mix together information from entries in $f$ and $f'$ to form new
entries.

@@

At this point I started thinking that my definitions of $\sqsubseteq$,
consistency, and $\sqcup$ were too simple, and that I needed to
incorporate ideas from the literature on filter models and
intersection types. As I've written about previously, my graph model
corresponds to a particular intersection type system, and perhaps a
different intersection type system would do the job. Recall that the
correspondence goes as follows: values correspond to types,
$\sqsubseteq$ corresponds to subtyping $<:$ (in reverse), and $\sqcup$
corresponds to intersection $\sqcap$. The various intersection type
systems primarily differ in their definitions of subtyping.  Given the
above proof attempt, I figured that I would need usual
co/contra-variant rule for function types and also the following rule
for distributing intersections over function types.
\[
  (A\to B) \sqcap (A \to C) <: A \to (B \sqcap C)
\]
This distributivity rule enables the ``mixing'' of information from
two different entries.

So I defined types as follows:
\[
  A,B,C,D ::= n \mid A \to B \mid A \sqcap B
\]
and defined subtyping according to the BCD intersection type system.
\begin{gather*}
A <: A \qquad \frac{A <: B \quad B <: C}{A <: C} \\[2ex]
A \sqcap B <: A \qquad
A \sqcap B <: B \qquad
\frac{C <: A \quad C <: B}{C <: A \sqcap B} \\[2ex]
\frac{C <: A \quad B <: D}{A \to B <: C \to D}
\qquad
(A\to B) \sqcap (A \to C) <: A \to (B \sqcap C)
\end{gather*}
I then adapted the definition of consistency to work over types.
(Because this definition uses negation, it is easier to define
consistency as a recursive function in Isabelle instead of as an
inductively defined relation.)
\begin{align*}
 n \sim n' &= (n = n') \\
 n \sim (C \to D) &= \mathit{false} \\
 n \sim (C \sqcap D) &= n \sim C \text{ and } n \sim D \\
 (A \to B) \sim n' &= \mathit{false} \\
 (A \to B) \sim (C \to D) &= 
    (A \sim C \text{ and } B \sim D) \text{ or } A \not\sim C \\
 (A \to B) \sim (C \sqcap D) &=
    (A \to B) \sim C \text{ and } (A \to B) \sim D \\
 (A \sqcap B) \sim n' &= A \sim n' \text{ and } B \sim n' \\
 (A \sqcap B) \sim (C \sqcap D) &= 
    A \sim C \text{ and } A \sim D \text{ and } 
    B \sim C \text{ and } B \sim D
\end{align*}

Turning back to the Join Theorem, I restated it in terms of the
intersection type system and rebranded it the Meet Theorem. Instead of
using the letter $\rho$ for environments, we shall switch to $\Gamma$
because they now contain types instead of values. \\

@@

For a long time I've disliked definition of subtyping in which
transitivity is given as a rule instead of proved as a theorem.  There
are several reasons for this: a subtyping algorithm can't directly
implement a transitivity rule (or any rule that is not syntax
directed), reasoning by induction or cases (inversion) is more
difficult, and it is redundant. Furthermore, the presense of the
transitivity rule means that subtyping does not satisfy the subformula
property. This term subformula property comes from logic, and means
that a derivation (proof) of a formula only mentions propositions that
are a part of the formulate to be proved. The transitivity rule breaks
this property because the type $B$ comes out of nowhere, it is not
part of $A$ or $C$, the types in the conclusion of the rule.

So I removed the transitivity rule and tried to prove transitivity.
For most type systems, proving the transitivity of subtyping is
straightforward. But what I soon realized that the addition of the
distributivity rule makes it significantly more difficult.  After
trying and failing to prove transitivity for some time, I resorted to
reading the literature. Unfortunately, it turns out that none of the
published intersection type systems satisfied the subformula property
and vast majority of them included the transitivity rule.  However,
there was one paper that offered some hope.  In a 2012 article in
Fundamenta Informaticae titled \emph{Intersection Types with Subtyping
  by Means of Cut Elimination}, Olivier Laurent defined subtyping
without transitivity and instead proved it, but his system still did
not satisfy the sub-formula property because of an addition rule for
function types. Nevertheless, Olivier indicated that he was interested
in finding a version of the system that did, writing
\begin{quote}
``it would be much nicer and much more natural to go through a
  sub-formula property''
\end{quote}

A lot of progress can happen in six years, so I sent an email to
Olivier. He replied,
\begin{quote}
``Indeed! I now have two different sequent-calculus systems which are
  equivalent to BCD subtyping and satisfy the sub-formula property.  I
  am currently writting a paper on this but it is not ready yet.''
\end{quote}
and he attached the paper draft and the Coq mechanization. What great
timing!  Furthermore, Olivier would be presenting the paper, titled
\emph{Intersection Subtyping with Constructors}, at the Workshop on
Intersection Types and Related System in Oxford on July 8, part of the
Federated Logic Conference (FLOC). I was planning to attend FLOC
anyways, for the DOMAINS workshop to celabrate Dana Scott's 85th
birthday.

@@

%%  LocalWords:  denotationally ctx subexpression eq denot lrcl env

>>>>

Many of the challenges revolved around the definitions of
$\sqsubseteq$, consistency, and $\sqcup$. Given that I already had
definitions for $\sqsubseteq$ and consistency, the obvious thing to
try was to define $\sqcup$ such that it would be the least upper bound
of $\sqsubseteq$. So I arrived at this partial function:
\begin{align*}
  n \sqcup n &= n \\
  f_1 \sqcup f_2 &= f_1 \cup f_2
\end{align*}
Now suppose we prove the Join Theorem by induction on $e$ and consider
the case for application: $e = (e_1 \app e_2)$. From $v \in \ESEM{e_1
  \app e_2}$ and $v' \in \ESEM{e_1 \app e_2}$ we have
\begin{itemize}
\item $f \in \ESEM{e_1}\rho$, $v_2 \in \ESEM{e_2}\rho$,  
  $v_3 \mapsto v_4 \in f$, $v_3 \sqsubseteq v_2$, and $v \sqsubseteq v_4$
  for some $f, v_2, v_3, v_4$.
\item $f' \in \ESEM{e_2}\rho'$, $v'_2 \in \ESEM{e_2}\rho'$,   
  $v'_3 \mapsto v'_4 \in f$, $v'_3 \sqsubseteq v'_2$, and $v' \sqsubseteq v'_4$
  for some $f', v'_2, v'_3, v'_4$.
\end{itemize}
By the induction hypothesis we have $f \sqcup f' \in \ESEM{e_1}$
and $v_2 \sqcup v'_2 \in \ESEM{e_2}$.
We need to show that 
\[
   v''_3 \mapsto v''_4 \in f \sqcup f' 
   \qquad
   v''_3 \sqsubseteq v_2 \sqcup v'_2
   \qquad
   v \sqcup v' \sqsubseteq v''_4
\]
But here we have a problem. Given our definition of $\sqcup$ in terms
of set union, there won't necessarily be a single entry in $f \sqcup
f'$ that combines the information from both $v_3 \mapsto v_4$ and
$v'_3 \mapsto v'_4$. After all, $f \sqcup f'$ contains all the entries
of $f$ and all the entries of $f'$, but the set union operation does
not mix together information from entries in $f$ and $f'$ to form new
entries.

@@

At this point I started thinking that my definitions of $\sqsubseteq$,
consistency, and $\sqcup$ were too simple, and that I needed to
incorporate ideas from the literature on filter models and
intersection types. As I've written about previously, my graph model
corresponds to a particular intersection type system, and perhaps a
different intersection type system would do the job. Recall that the
correspondence goes as follows: values correspond to types,
$\sqsubseteq$ corresponds to subtyping $<:$ (in reverse), and $\sqcup$
corresponds to intersection $\sqcap$. The various intersection type
systems primarily differ in their definitions of subtyping.  Given the
above proof attempt, I figured that I would need the usual
co/contra-variant rule for function types and also the following rule
for distributing intersections over function types.
\[
  (A\to B) \sqcap (A \to C) <: A \to (B \sqcap C)
\]
This distributivity rule enables the ``mixing'' of information from
two different entries.

So I defined types as follows:
\[
  A,B,C,D ::= n \mid A \to B \mid A \sqcap B
\]
and defined subtyping according to the BCD intersection type system
(\emph{Lambda Calculus with Types}, Barendregt et al. 2013).
\begin{gather*}
A <: A \qquad \frac{A <: B \quad B <: C}{A <: C} \\[2ex]
A \sqcap B <: A \qquad
A \sqcap B <: B \qquad
\frac{C <: A \quad C <: B}{C <: A \sqcap B} \\[2ex]
\frac{C <: A \quad B <: D}{A \to B <: C \to D}
\qquad
(A\to B) \sqcap (A \to C) <: A \to (B \sqcap C)
\end{gather*}
I then adapted the definition of consistency to work over types.
(Because this definition uses negation, it is easier to define
consistency as a recursive function in Isabelle instead of as an
inductively defined relation.)
\begin{align*}
 n \sim n' &= (n = n') \\
 n \sim (C \to D) &= \mathit{false} \\
 n \sim (C \sqcap D) &= n \sim C \text{ and } n \sim D \\
 (A \to B) \sim n' &= \mathit{false} \\
 (A \to B) \sim (C \to D) &= 
    (A \sim C \text{ and } B \sim D) \text{ or } A \not\sim C \\
 (A \to B) \sim (C \sqcap D) &=
    (A \to B) \sim C \text{ and } (A \to B) \sim D \\
 (A \sqcap B) \sim n' &= A \sim n' \text{ and } B \sim n' \\
 (A \sqcap B) \sim (C \sqcap D) &= 
    A \sim C \text{ and } A \sim D \text{ and } 
    B \sim C \text{ and } B \sim D
\end{align*}

Turning back to the Join Theorem, I restated it in terms of the
intersection type system and re-branded it the Meet Theorem. Instead of
using the letter $\rho$ for environments, we shall switch to $\Gamma$
because they now contain types instead of values. \\

@@

For a long time I've disliked definition of subtyping in which
transitivity is given as a rule instead of proved as a theorem.  There
are several reasons for this: a subtyping algorithm can't directly
implement a transitivity rule (or any rule that is not syntax
directed), reasoning by induction or cases (inversion) is more
difficult, and it is redundant. Furthermore, the presence of the
transitivity rule means that subtyping does not satisfy the sub-formula
property. This term sub-formula property comes from logic, and means
that a derivation (proof) of a formula only mentions propositions that
are a part of the formulate to be proved. The transitivity rule breaks
this property because the type $B$ comes out of nowhere, it is not
part of $A$ or $C$, the types in the conclusion of the rule.

So I removed the transitivity rule and tried to prove transitivity.
For most type systems, proving the transitivity of subtyping is
straightforward. But what I soon realized that the addition of the
distributivity rule makes it significantly more difficult.  After
trying and failing to prove transitivity for some time, I resorted to
reading the literature. Unfortunately, it turns out that none of the
published intersection type systems satisfied the sub-formula property
and vast majority of them included the transitivity rule.  However,
there was one paper that offered some hope.  In a 2012 article in
Fundamenta Informaticae titled \emph{Intersection Types with Subtyping
  by Means of Cut Elimination}, Olivier Laurent defined subtyping
without transitivity and instead proved it, but his system still did
not satisfy the sub-formula property because of an addition rule for
function types. Nevertheless, Olivier indicated that he was interested
in finding a version of the system that did, writing
\begin{quote}
``it would be much nicer and much more natural to go through a
  sub-formula property''
\end{quote}

A lot of progress can happen in six years, so I sent an email to
Olivier. He replied,
\begin{quote}
``Indeed! I now have two different sequent-calculus systems which are
  equivalent to BCD subtyping and satisfy the sub-formula property.  I
  am currently writing a paper on this but it is not ready yet.''
\end{quote}
and he attached the paper draft and the Coq mechanization. What great
timing!  Furthermore, Olivier would be presenting the paper, titled
\emph{Intersection Subtyping with Constructors}, at the Workshop on
Intersection Types and Related System in Oxford on July 8, part of the
Federated Logic Conference (FLOC). I was planning to attend FLOC
anyways, for the DOMAINS workshop to celabrate Dana Scott's 85th
birthday.

@@

%%  LocalWords:  denotationally ctx subexpression eq denot lrcl env
%%  LocalWords:  iff Hah pdf subtyping contra distributivity BCD et
%%  LocalWords:  Barendregt al Fundamenta Informaticae sequent
